### 2014-09-05 text/Chapter05.tex
Substitution for structural coeffects (Chapter 5)

Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
different notion of context. As before, the properties of the flat coeffect algebra guarantee that 
certain equational properties hold for all instances of the calculus.

\subsection{From flat to structural coeffects}

The discussion of equational theory for flat calculus in Section~\ref{sec:flat-syntax} considered
different reductions that can be used as a basis for operational semantics for calculi satisfying
different additional properties, as none of the techniques works universally. The structural 
coeffect calculus has more desirable equational properties -- in particular, both $\beta$-reduction
and $\eta$-expansion can be proved using just the properties of strucutral coeffect algebra. Thus
this section focuses on these reductions. 
Using the terminology of Pfenning and Davies \cite{logic-modal-reconstruction},
the structural coeffect calculus satisfies both are \emph{local soundness} and \emph{local completeness}.

% The structural coeffect calculus clstr~associates information with individual variables. This means
% that when an expression requires certain context, we know from what scope it comes -- the
% context must be provided by a scope that defines the associated variable, which is either a
% lambda abstraction or global scope. This distinguishes the structural system from the flat system
% where context could have been provided by any scope and the lambda rule allowed arbitrary 
% splitting of context requirements between the two scopes (or declaration and caller site).

\subsection{Holes substitution lemma}

if

\begin{equation*}
\coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s
\end{equation*}

and

\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_s}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
  \coctx{x_k\!:\!\tau_s}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
\end{equation*}

then

\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{r}}} \,|\, \ldots \,|\, 
  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{r}}}\,] \vdash \subst{\subst{e_r}{x_1}{e_s}\ldots}{x_k}{e_s} : \tau_r
\end{equation*}

\subsection{Structural coeffect systems}
\label{prop:structural}

For structural coeffect systems, recall that coeffects are vectors
with $\czip = \cpar = \cons$ (vector concatenation) and $\czero =
\langle{}\rangle$ (the empty vector), thus coeffect annotations
comprise the \emph{free monoid} \ie, lists over the coeffect scalars
(although we continue using the vector terminology). We first show
substitution:

\begin{lemma}[Substitution lemma] 
\label{thm:structural-subst}
In a structural coeffect calculus with a coeffect scalar structure $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$: 
\[
\begin{array}{l}
\coctx{\Gamma}{\aclrd{S}} \vdash e_s : \sigma  
\;\;  \wedge  \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{\aclrd{R_1} \cons\; \alift{\cclrd{r}} \cons\; \aclrd{R_2} } \vdash e_r : \tau\\
\quad \Rightarrow \; \coctx{\Gamma_1,\Gamma,\Gamma_2}{\aclrd{R_1} \cons\; (\cclrd{r} \,\aseq\, \aclrd{S}) \cons\; \aclrd{R_2}} \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\label{lemm:structural-subst}
\end{lemma}
\begin{proof}
By induction over the $e_2$ derivation using the free monoid structure ($\C, \cons, \langle{}\rangle$)
and coeffect scalar axioms (full proof~\cite{appendix}). %Full proof in the Appendix.
\end{proof}

\noindent
Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
Using the substitution lemma, we can demonstrate $\beta$-equality:
%
\begin{equation*}
\inference{
  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
\equiv
\subst{e_1}{x}{e_2} : \tau}
\end{equation*}
%%
\noindent
As a result, $\beta$-reduction preserves the type and coeffects of a term.
This gives the following subject reduction property:

@@

% ==================================================================================================
%
%                               ##      
%      ####                     ##      
%        ##                     ##      
%        ##   ##   ##  ## ###   ##  ##  
%        ##   ##   ##  ###  ##  ## ##   
%    ##  ##   ##   ##  ##   ##  ####    
%    ##  ##   ##  ###  ##   ##  ## ##   
%     ####     ### ##  ##   ##  ##  ##  
%                                      
% ==================================================================================================

\newcommand{\ctimes}{ \hat{\times} }
\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }

\newcommand{\csprd}[0]{ \times }
\newcommand{\csvop}[0]{ \vee }
\newcommand{\cmerge}[0]{ \cpar }
\newcommand{\csplit}[0]{ \czip }

\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}
\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}

\paragraph{Internalized substitution.}
Before looking at properties of the evaluation, we consider let binding, which can be 
viewed as internalized substitution. The typing rule \rulest{let} can be derived from 
application and abstraction as follows.

\begin{lemma}[Definition of let binding]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
\end{lemma}
\begin{proof}
The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
correspond with the typing rule \rulest{let}:
\begin{equation*}
\inference[]
  {\inference
    {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
\end{equation*}
\end{proof}

The term $e_2$ which is substituted in $e_1$ is checked in a different variable
and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
the variable that we are substituting for. The let binding captures substitution for a 
specific variable (the context is of a form $\ctyp{r \csprd s}{\Gamma, v:\tau}$).
For a general substitution, we need to define the notion of context with a hole.

\newcommand{\chole}[4]{\ctyp{#1}{#2}[#3|#4]}
\paragraph{Substitution and holes.}
In \clstr, the structure of the variable context is not a set, but a tree. When substituting
for a variable, we need to replace the variable in the context with the context of the
substituted expression. In general, this can occur anywhere in the tree. To formulate the
statement, we define contexts with holes, written $\Delta[-]$. Note that there is a hole
in the free variable context and in a corresponding part of the coeffect tag:

\begin{equation*}
\begin{array}{rcl}
\Delta[-] &::=& \ctyp{1}{()} \\
       & \sep & \ctyp{r}{(x:\tau)}\\
       & \sep & \ctyp{-}{(-)} \\
       & \sep & \ctyp{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
\end{array}
\end{equation*}
%
Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
general substitution lemma for \clstr.

\begin{figure}

\begin{equation*}
\begin{array}{rcl}
                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
\chole{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \csprd r_2'}{(\Gamma_1', \Gamma_2')}\\
&\multicolumn{2}{l} {
  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
}
\end{array}
\end{equation*}
\caption{The definition of hole filling operation for $\Delta[-]$}
\label{fig:structural-hole-fill}
\vspace{-2em}
\end{figure}

\begin{lemma}[Substitution Lemma]
\label{thm:structural-subst}
If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
then $\chole{r}{\Gamma}{R \csvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
\end{lemma} 
\begin{proof}
Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
$(S, \csvop, 0, \csprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
\end{proof}

\begin{theorem}[Subject reduction]
\label{thm:structural-reduction}
If $\ctyp{r}{\Gamma} \vdash e_1 : \tau$ and $e_1 \twoheadrightarrow_\beta e_2$ then 
$\ctyp{r}{\Gamma} \vdash e_2 : \tau$.
\end{theorem}
\begin{proof}
Direct consequence of Lemma~\ref{thm:structural-subst} (see Appendix~\ref{app:structural-reduction}).
\end{proof}

\paragraph{Local soundness and completeness.}
As with the previous calculus, we want to guarantee that the introduction and
elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
This can be done by showing \emph{local soundness} and \emph{local completeness},
which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
case of subject reduction and the latter is proved by a simple derivation:
%
\begin{theorem}[Local soundness]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
\end{theorem}
\begin{proof}
Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
\end{proof}
%
\begin{theorem}[Local completeness]
If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
\end{theorem}
\begin{proof} The property is proved by the following typing derivation:
%
\begin{equation*}
\inference
  {\inference
     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
     {\ctyp{r \csprd (s \csvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\qedhere
\end{equation*}
\end{proof}
%
In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
guarantees that $s\csvop0 = s$. Recall that in \clflt, the typing derivation for
$\lambda x.f x$ required for local completeness was not the only possible derivation.
In the last step, it was possible to split the coeffect tag arbitrarily between the 
context and the function type.

In the \clstr~calculus, this is not, in general, the case. The $\csprd$ operator is not required 
to be associative and to have units and so a unique splitting may exist.  For example, if we
define $\csprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
$t$, there are unique $r$ and $s$ such that $t = r \csprd s$. However, if the $\csprd$ operation 
has additional properties, then there may be other possible derivation.

% ==================================================================================================

\section{Examples of structural coeffects}

\subsection{Example: Liveness analysis}
\label{sec:structural-neededness}

\subsection{Example: Data-flow (revisited) }
\label{sec:structural-tainting}

\todo{Also, consider additional language features that we consider for flat coeffects
  (mainly recursion and possibly conditionals)}

>>>>

Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
different notion of context and thus a different operational interpretation. As before, the 
properties of the flat coeffect algebra guarantee that certain equational properties hold for 
all instances of the calculus.

We start the discussion by briefly considering the key aspects
that make the equational theory of flat and structural coeffects different.

\subsection{From flat coeffects to structural coeffects}
\label{sec:struct-syntax-intro}

When discussing equational theory for the flat calculus in Section~\ref{sec:flat-syntax}, we 
noted that no single technique works universally for all flat coeffect calculi. We considered
multiple different reductions that can be used as the basis for operational semantics for
calculi satisfying different additional properties.

The structural coeffect calculus has more desirable equational properties. In particular, 
we can prove both $\beta$-reduction and $\eta$-expansion using just the properties of strucutral 
coeffect algebra. For this reason, we focus on these two reductions in this section.
Using the terminology of Pfenning and Davies \cite{logic-modal-reconstruction},
the structural coeffect calculus satisfies both the \emph{local soundness} and 
the \emph{local completeness} properties.

\paragraph{Substitution for flat coeffects.} 
The more interesting variant of the substitution lemma for flat coeffects 
(Lemma~\ref{thm:cbn-substitution-bot}) required all operations of the flat coeffect algebra
to coincide. This enables the substitution to preserve the type of expressions, because all
additional requirements arising as the result of the substitution can be associated with the
declaration context. For example, consider the following example where implicit parameter 
$\ident{?offset}$ is substituted for the variable $y$:
%
\begin{equation*}
\begin{array}{rclcll}
 \coctx{y\!:\!\ident{int}}{\cclrd{\emptyset}} &\narrow{\vdash}& \lambda x.y + \ident{?total} &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{before}) \\
\coctx{()}{\cclrd{ \{ \ident{?offset} \} }} &\narrow{\vdash}& \lambda x.\ident{?offset} + \ident{?total} &\narrow{:}&
  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{after})
\end{array}  
\end{equation*}
%
The typing judgement obtained in (\emph{after}) preserves the type of the expression
(function value) from the original typing (\emph{before}). This is possible thanks to the
non-determinism involved in lambda abstraction -- as all operators of the flat coeffect
algebra used here are $\cup$, we can place the additional requirement on the outer context.
Note that this is not the \emph{only} possible typing, but it is \emph{permissible} typing.

Here, the flat coeffect calculus gives us typing with limited \emph{precision}, but 
enough \emph{flexibility} to prove the substitution lemma.

\paragraph{Substitution for structural coeffects.}
In contrast, the substitution lemma for structural coeffects can be proven ecause 
structural coeffect systems provide enough \emph{precision} to identify exactly with 
which variable should a context requirement be associated.

The following example shows a situation similar to the previous one -- here, we use
structural data-flow calculus (with the \kvd{prev} construct to obtain previous
value of an expression) and we substitute $w+z$ for $y$:
%
\begin{equation*}
\begin{array}{rclcll}
 \coctx{y\!:\!\ident{int}}{ \alift{\cclrd{2}} } 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{before}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\cclrd{\ast}\alift{\cclrd{1},\cclrd{1}}} 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{after}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}} 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{final})
\end{array}  
\end{equation*}
%
The type of the function does not change, because the structural type system associates
the annotation $\cclrd{1}$ with the bound variable $x$ and the substitution does not 
affect how the variable $x$ is used.

The other aspect demonstrated in the example is how the coeffect of the substituted 
variable affects the free-variable context of the substituted expression. Here, the
original variable $y$ is annotated with $\cclrd{2}$ and we substitute it for an
expression $w+z$ with free variables $w,z$ annotated with $\alift{\cclrd{1}, \cclrd{1}}$.
The substitution applies the operation $\aseq$ (modelling sequential composition) to 
the annotation of the new context -- in the above example
$\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}=\alift{\cclrd{2}, \cclrd{2}}$.

\subsection{Holes and substitution lemma}

As demonstrated in the previous section, reduction (and substitution) in the structural coeffect 
calculus may need to replace a \emph{single} variable with a \emph{vector} of variables. 
Furthermore, we may also need to substitute for multiple variables in the variable context.

For example, the expression $\lambda x.x+x$ is type-checked by type-checking $x_1 + x_2$,
contracting $x_1$ and $x_2$ and then applying lambda abstraction. During the reduction 
$(\lambda x.x+x)~(y+z)$ we thus need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
This is similar to substitution lemma in other structural variants of $\lambda$-calculus,
for example in the bunched typing system \cite{substruct-bunched}.
To express the substitution lemma, we define the notion of a \emph{context with holes}

\begin{definition}[Context with holes]
A \emph{context with holes} is a context such as $\coctx{x_1\!:\!\tau_1, \ldots, x_k\!:\!\tau_k}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_k}}}$,
where some of the variable typings $x_i\!:\!\tau_i$ and corresponding coeffects $\cclrd{r_i}$ 
are replaced by \emph{holes} written as $-$.

\begin{equation*}
~\Delta[\,\coctx{-}{-}\,]_n = \Delta[\,\underbrace{\coctx{-}{-} \,|\, \ldots \,|\, \coctx{-}{-}}_{ n-\textit{times} } \,]
\end{equation*}
\begin{equation*}
\begin{array}{lcl}
 \Delta[\,\coctx{-}{-}\,]_n &\narrow{:=}&
   \coctx{-, \Gamma}{\alift{-} \atimes \aclrd{\textbf{s}} }
   \qquad \textnormal{where}~\coctx{\Gamma}{\aclrd{\textbf{s}}} \in \Delta[\,\coctx{-}{-}\,]_{n-1} \\
 \Delta[\,\coctx{-}{-}\,]_n &\narrow{:=}&
   \coctx{x\!:\!\tau, \Gamma}{\alift{\cclrd{r}} \atimes \aclrd{\textbf{s}} }
   \qquad \textnormal{where}~\coctx{\Gamma}{\aclrd{\textbf{s}}} \in \Delta[\,\coctx{-}{-}\,]_n \\
 \Delta[\,\coctx{-}{-}\,]_0 &\narrow{:=}&
   \coctx{()}{\alift{}}
\end{array} 
\end{equation*} 
\end{definition}

\noindent
A context with $n$ holes may either start with a hole, followed by a context with $n-1$ holes, or it may
start with a variable followed by a context with $n$ holes. Note that the definition ensures that
the location of variable holes correspond to the locations of coeffect annotation holes. Given a context
with holes, we can fill the holes with other contexts using the \emph{hole filling} operation and
obtain an ordinary coeffect-annotated context.

\begin{definition}[Hole filling] Given a context with $n$ holes $\coctx{\Delta}{\textbf{s}} \in 
\Delta[\,\coctx{-}{-}\,]_n$, the hole filling operation written as 
$\coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\mathbf{r_1}}} \,|\, \ldots \,|\, \coctx{\Gamma_n}{\aclrd{\mathbf{r_n}}} \,]$,
which replaces the holes by the specified variables and corresponding coeffect annotations, is defined as:
%
\begin{equation*}
\begin{array}{rcl}
 \coctx{-, \Delta}{\alift{-} \atimes \textbf{s} }
    ~[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] &\narrow{=}&
      \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}_1} \atimes \aclrd{\textbf{r}_2} } \\
 && \qquad \hspace{-9.3em} \textnormal{where}~\coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} =
     \coctx{\Delta}{\textbf{s}}[\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] 
\\[0.5em]
 \coctx{x_1\!:\!\tau, \Delta}{\alift{\cclrd{r_1}} \atimes \textbf{s} }
    ~[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] &\narrow{=}&
      \coctx{x_1\!:\!\tau, \Gamma_2}{\alift{\cclrd{r_1}} \atimes \aclrd{\textbf{r}_2} } \\
 && \qquad \hspace{-9.3em} \textnormal{where}~\coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} =
     \coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] 
\\[0.5em]
 \coctx{()}{\alift{}}~[~] &\narrow{=}& \coctx{()}{\alift{}}
\end{array} 
\end{equation*} 
\end{definition}

\noindent
When we substitute an expression with coeffects $\aclrd{\textbf{t}}$ (associated with variables $\Gamma$)
for a variable that has coeffects $\cclrd{s}$, the resulting coeffects of $\Gamma$ need to 
combine $\aclrd{\textbf{t}}$ and $\cclrd{s}$. Unlike in the flat coeffect systems, the structural 
substitution does not require all coeffect algebra operations to coincide and so the combination
is more interesting than in the bottom-pointed substitution for flat coeffects, where it used
the only available operator (Lemma~\ref{thm:cbn-substitution-bot}).

Substitution can be seen as sequential composition. Informally -- we first need to obtain the 
value of the expression (requiring $\aclrd{\textbf{t}}$) and then use it in context with 
requirements $\cclrd{s}$. Thus the free variables of the expression \emph{after} substitution
are annotated with $\cclrd{s} \aseq \aclrd{\textbf{t}}$, using the (scalar-vector extension)
of the sequential composition operator $\cseq$.

\begin{lemma}[Multi-nary substitution]
Given an expression with multiple holes filled by variables $x_i\!:\!\tau_i$ with coeffects $\cclrd{s_k}$:
%
\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_1}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
  \coctx{x_k\!:\!\tau_k}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
\end{equation*}
%
and a expressions $e_i$ with free-variable contexts $\Gamma_i$ annotated with $\aclrd{\mathbf{t_i}}$:
%
\begin{equation*}
\coctx{\Gamma_1}{\aclrd{\textbf{t}_1}} \vdash e_1 : \tau_1 
\quad \ldots \quad
\coctx{\Gamma_k}{\aclrd{\textbf{t}_k}} \vdash e_k : \tau_k
\end{equation*}
%
then substituting the expressions $e_i$ for variables $x_i$ results in an expression with a context
where the original holes are filled by contexts $\Gamma_i$ with coeffects $\cclrd{s_i} \aseq \aclrd{\mathbf{t_i}}$:
%
\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{t}_1}} \,|\, \ldots \,|\, 
  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{t}_k}}\,] \vdash \subst{\subst{e_r}{x_1}{e_1}\ldots}{x_k}{e_k} : \tau_r
\end{equation*}
\end{lemma}
\begin{proof}
By induction over $\vdash$, using the multi-nary aspect of the substitution 
in the proof of the contraction case. See Appendix~?.
\end{proof}

\subsection{Reduction and expansion}

\todo{Exterminate!}

~

Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
Using the substitution lemma, we can demonstrate $\beta$-equality:
%
\begin{equation*}
\inference{
  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
\equiv
\subst{e_1}{x}{e_2} : \tau}
\end{equation*}
%%
\noindent
As a result, $\beta$-reduction preserves the type and coeffects of a term.
This gives the following subject reduction property:

@@
